Choosing Color Palettes for Statistical Graphics

### 2006-09-29 trunk/vcd/paper/Color/draft.Rnw
minor changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@577 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

Due to these three color axes, colors are typically described as
locations in a 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces are defined by three dimensions
that try to capture
\begin{enumerate}
  \item \textbf{hue} (dominant wavelength)
  \item \textbf{chroma} (colorfulness)
  \item \textbf{luminance} (brightness, `amount of gray')
\end{enumerate}

@@

@@

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ and different combinations of maximal chroma (0, 80 and 100, respecitvely) and
minimal luminance (30, 30 and 50, respectively). The first two combinations are also
shown in the first two rows of Figure~\ref{fig:pal-s-examples}. The right panel of
Figure~\ref{fig:pal-s-construct} shows that the exact same strategy is not possible 
for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L = 75$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is show for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

@@

Given sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---with slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
These particular hues were chosen because they have rather similar
chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
that are sufficiently similar in their chroma/luminance plane to allow for the same
luminance and/or chroma contrasts.

@@

In this section, we show a few examples for the various types of palettes applied
to statistical graphics. The first illustration is visualization of a bivariate
density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
shows heatmaps of a bivariate kernel density estimate of waiting times between
and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.

@@

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU better in the South; furthermore, the
Die Linke performed particularly well in the Eastern provinces and the Saarland.

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlation (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group which are
shown to be significant by the shading based on the Pearson residuals. Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.

>>>>

%%FIXME: what to say about ColorBrewer

@@

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces are defined by three dimensions
that try to capture
\begin{enumerate}
  \item \textbf{hue} (dominant wavelength)
  \item \textbf{chroma} (colorfulness)
  \item \textbf{luminance} (brightness, amount of light)
\end{enumerate}

@@

%%FIXME: briefly explain interface

@@

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

@@

Given sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---with slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
These particular hues were chosen because they have rather similar
chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
%%FIXME: symmetric difference
As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
that are sufficiently similar in their chroma/luminance plane to allow for the same
luminance and/or chroma contrasts.

@@

In this section, we show a few examples for the various types of palettes applied
to statistical graphics. The first illustration is a visualization of a bivariate
density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
shows heatmaps of a bivariate kernel density estimate of waiting times between
and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.

@@

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and the Saarland.

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group which are
shown to be significant by the shading based on the Pearson residuals. Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.
